Climate Change Adaptation: Uncertainty Associated with Sequential Finite Period Decisions under Nonstationarity

### 2018-02-16 writeup/Outline/main.tex
plot the stationary and trend sequences

We generate synthetic flood sequences from a simple conceptual model where the expected flood varies as a function of a single climate index plus time following the log-normal distribution:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}
The parameter \( \mu(t) \) is generated using two different approaches, summarized below, while for both we assume a constant coefficient of variation:
\begin{equation}
  \sigma(t) = \begin{cases} \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ \sigma_\text{min} & \qqtext{else} \end{cases}
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.
Though both approaches used to generate sequences \( \mu(t) \) use ighly simplified models neglecting many important mechanisms, they enable us to generate many synthetic sequences of \(\qty[M + N]\) years which exhibit chaotic but organized quasi-periodic behavior (regime behavior or low-frequency variability) plus a trend term, thereby capturing many essential features of hydrological variability.

In \cref{sec:sequence-realistic} we show sequences of streamflow generated by this model for some choices of parameters and discuss the processes which they can reasonably represent.

@@

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a quasi-oscillatory manner with variability on the order of \SIrange{4}{6}{year} modulated by low-frequency oscillation \citep[see][for a comprehensive review]{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citep{Ramesh2017}.
To create an annual time series, we average the October-December months to create a time series \( x(t) \) where \( t \) is taken in annual steps \(1, \ldots, \num{100000} \).
The parameters of the distribution vary in time according to the sampled NINO3 index denoted \( x(t) \) as well as time itself, representing secular trends.
First a starting point in the \SI{100000}{year} ENSO was randomly chosen, then \(\mu\) is determined by
\begin{equation}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.

\subsubsection{Markov Chain}

Alternatively, we consider a 2-state Markov chain with a constant transition probability which is the same for both states, such that the transition matrix is
\begin{equation*}
  P = \mqty[\pi & 1-\pi \\ 1-\pi & \pi]
\end{equation*}
This is first used to generate a sequence of states \(S(t)\).
The value \( \mu(t) \) depends only on the \(S(t)\) and the time itself:
\begin{equation*}
  \mu(t) = \begin{cases}
    \mu_{0,1} + \gamma_1 \qty(t - t_0) & \qqtext{if} S(t) = 1 \\
    \mu_{0,2} + \gamma_2 \qty(t - t_0) & \qqtext{if} S(t) = 2
  \end{cases}
\end{equation*}
For simplicity, we assume that the coefficient of variation \( \alpha \) is the same for both states.

@@

Because we do not, in general, know the true distribution of annual-maximum floods, we compare a stationary and non-stationary parameterizations of the generalized extreme value (GEV) statistical distribution (eq.~\ref{eq:GEV}) for estimating the distribution of \(Q(t)\) for the future \(M\)-year period.
\begin{equation}
  \label{eq:GEV}
  f(x \big| \mu, \sigma, \xi) = \frac{1}{\sigma}
    \begin{cases}
      \qty[1 + \xi \frac{x - \mu}{\sigma}] ^ {(-\flatfrac{1}{\xi})-1} \exp \qty[-\qty(1+\xi \frac{x - \mu}{\sigma})^{\flatfrac{-1}{\xi}}], & \xi \neq 0 \\
      \exp \qty[-\frac{x-\mu}{\sigma}] \exp \qty[- \exp \qty(- \frac{x-\mu}{\sigma})], & \xi = 0
    \end{cases}
\end{equation}
Where \(\mu{}\) is the location parameter, \(\sigma{}\) is the shape parameter, and \(\xi{}\) is the scale parameter of the GEV distribution.
For each simulation, the first \(N\) years \(\qty(t=[t_0-N+1, \ldots, t_0])\) are treated as observations.
Then, once the appropriate time parameterization is selected (\cref{tab:model-fitting}), the parameters are fit in a Bayesian framework to fully represent the posterior uncertainty.
Computation is carried out in the \texttt{python} environment, with simulation using the \texttt{numpy} package \citep{vanderWalt2011} and fitting using the No-U-Turn Sampler (NUTS) \citep{Hoffman2014} as implemented in the \texttt{stan} probabilistic programming language \citep{Carpenter2016}.
Weak (uninformative) priors are chosen based on recommendations from \citet{Martins2000}; the exact parameterizations are described in following sections which describe specific experiments.

@@

As \cref{sec:method-stationary} except that
\begin{equation*}
  \mu = \mu_0 + \beta_\mu \qty(t - t_0)
\end{equation*}
and so the parameters \( \mu_0, \beta_\mu \) must be estimated rather than just \(\mu\).

@@

\subsection{Flood Estimation and Uncertainty\label{sec:methods-uncertainty}}

Finally, we consider a contract in which an insurer agrees to pay the entire coverage cost \(C\) every time that the trigger threshold \(Q_T\) is exceeded.
Although \(Q_T\) is in general a decision variable, we treat it as fixed.
Then, using the posterior distribution of the parameters of the chosen statistical model (from \cref{tab:model-fitting}) we estimate the probability that \(Q(t) \geq Q_T\) for \(t \in \vb{t}\), which is equivalent to \(p_T(t)\).

For a particular year, the fair price for the insurance premium is \(\expectation \qty(p_T) C\).
However, there is typically also a risk premium associated with the policy that prices the uncertainty in this estimate (since the trigger \(Q_T\) is specified in the contract and thus not uncertain).
This risk premium is proportional to the variance of \(p_T\), or more generally to its uncertainty distribution.
The insurance premium for a particular year is thus
\begin{equation} \label{eq:premium}
  Z(t) = C \qty[\expectation \qty(p_T(t)) + R \variance \qty(p_T(t))]
\end{equation}
where \(R \geq 0\) is a coefficient describing the cost of tue uncertainty in the estimate of \(p_T(t)\).

In this paper we consider the price of a contract of \(M\) years to be simply the summed price of the contract for each individual year.
We assume that the discount rate is equal to inflation (\ie{} the inflation-adjusted discount rate is zero) for simplicity.
A more comprehensive pricing mechanism for the financial instrument could incorporate more or different information about the uncertainty in the estimate, and should consider the price of the \(M\) years as a portfolio (\ie{} should consider correlation between the different years).
Such an approach, however, is not necessary for this conceptual study.

@@

\begin{figure}
  \includegraphics[width=\textwidth]{../../figs/stationary_sequences.pdf}
  \caption{50 simulated sequences created from the Cane-Zebiak model  \(N=50, M=50, t_0=0, \mu_0=6.5, \beta_\mu=0.5, \gamma=0, \alpha=0.075, \sigma_\text{min}=0.01\). 49 of the sequences are shown in gray and a single randomly chosen sequence is highlighted in blue for emphasis. (L) shows the time series of log streamflow as a function of year, while (R) shows the distribution of 500 sequences over the historical and future periods.}\label{fig:stationary-sequences}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{../../figs/trend_sequences.pdf}
  \caption{As \cref{fig:stationary-sequences} except \(\gamma=0.01\).}
\end{figure}

@@

@@

@@

We still need to find a good case study -- it can be a highly simplified one.
We can come up with some numbers and a single streamflow sequence that has been observed.

@@

@@

\section*{Notation}

@@

% -----------------------------------------------------------------------------
% END HERE
% -----------------------------------------------------------------------------
\clearpage
\printbibliography{}

>>>>

We generate synthetic flood sequences from two different simple conceptual models that represent chaotic dynamics and long memory.

For both cases, the expected value of the annual maximum flood flood follows a log-normal distribution:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}
where the parameters \( \mu(t), \sigma(t) \) are allowed to vary in time following the two approaches outlined below.
For simplicitiy we assume a constant coefficient of variation:
\begin{equation}
  \sigma(t) = \begin{cases} \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ \sigma_\text{min} & \qqtext{else} \end{cases}
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.

Though both approaches used to generate sequences \( \mu(t) \) useh highly simplified models neglecting many important mechanisms, they enable us to generate many synthetic sequences of \(\qty[M + N]\) years which exhibit chaotic but organized quasi-periodic behavior (regime behavior or low-frequency variability) plus a trend term, thereby capturing some essential features of hydrological variability.

In \cref{sec:sequence-realistic} we show sequences of streamflow generated by both models for some choices of parameters and discuss the processes which they can reasonably represent.

@@

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a quasi-oscillatory manner with variability on the order of \SIrange{4}{6}{year} modulated by low-frequency oscillation \citep[see][for a comprehensive review]{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citet{Ramesh2017}.
A \SI{1000}{year} subset of this data is shown in \cref{fig:enso-ts} and its spectrum in \cref{fig:enso-spectrum}.
To create an annual time series, we average the October-December months to create a time series \( x(t) \) where \( t \) is taken in annual steps \(1, \ldots, \num{100000} \).

First a sequence of length \(M+N\) is randomly chosen from the \SI{100000}{year} ENSO sequence.
Next, \(\mu\) is calculated following by
\begin{equation}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.

\subsubsection{Markov Chain\label{methods-markov}}

To create sequences of floods with strong memory and regime-like behavior, we use a 2-state Markov chain to sample a sequence of states that are used to generate \(\mu(t\).

This Markov chain is given a constant probability of persistence which is the same for both states, such that the transition matrix is
\begin{equation*}
  P = \mqty[\pi & 1-\pi \\ 1-\pi & \pi]
\end{equation*}
This is first used to generate a sequence of states \(S(t)\).
The value \( \mu(t) \) depends only on the \(S(t)\) and the time itself:
\begin{equation*}
  \mu(t) = \begin{cases}
    \mu_{0,1} + \gamma_1 \qty(t - t_0) & \qqtext{if} S(t) = 1 \\
    \mu_{0,2} + \gamma_2 \qty(t - t_0) & \qqtext{if} S(t) = 2
  \end{cases}
\end{equation*}
For simplicity, we assume that the coefficient of variation \( \alpha \) is the same for both states.
We also assume \( \mu_{0, 1} \gg \mu_{0, 2} \) so that state 1 can be interpreted as the ``wet'' state and state 2 as the ``dry'' state.

@@

We do not, in general, know the true distribution of annual-maximum floods, and there has been substantial discussion in the literature as to the choice of distribution to use for stationary flood frequency analysis.
A reasonable choice is the generalized extreme value (GEV) statistical distribution:
\begin{equation}
  \label{eq:GEV}
  f(x \big| \mu, \sigma, \xi) = \frac{1}{\sigma}
    \begin{cases}
      \qty[1 + \xi \frac{x - \mu}{\sigma}] ^ {(-\flatfrac{1}{\xi})-1} \exp \qty[-\qty(1+\xi \frac{x - \mu}{\sigma})^{\flatfrac{-1}{\xi}}], & \xi \neq 0 \\
      \exp \qty[-\frac{x-\mu}{\sigma}] \exp \qty[- \exp \qty(- \frac{x-\mu}{\sigma})], & \xi = 0.
    \end{cases}
\end{equation}
We will refer to \(\mu{}\) is the location parameter, \(\sigma{}\) as the shape parameter, and \( \xi{} \) as the parameter.

For each simulation, the first \(N\) years \(\qty(t=[t_0-N+1, \ldots, t_0])\) are treated as observations.
Then, once the appropriate time parameterization is selected (\cref{tab:model-fitting}), the parameters are fit in a Bayesian framework to fully represent the posterior uncertainty as implemented in the \texttt{stan} probabilistic programming language \citep{Carpenter2016}.
Weak (uninformative) priors are chosen based on recommendations from \citet{Martins2000}; the exact parameterizations are described in following sections which describe specific experiments.

@@

One approach for incorporating secular change into the GEV model is to add a time trend to the location parameter \citep[\ie{}][]{Bracken2017} such that
\begin{equation*}
  \mu = \mu(t) = \mu_0 + \beta_\mu \qty(t - t_0).
\end{equation*}
We fit the trend GEV model in exactly the same way as the stationary GEV model, except that the parameters  \( \mu_0, \beta_\mu \) must be estimated rather than just \( \mu \).

@@

\texttt{How much more do we need to say about HMM?}

\subsection{Reproducability}

All computation was carried out in the \texttt{python} environment, making particular use of the \texttt{xarray}, \texttt{numpy}, \texttt{pandas}, and \texttt{matplotlib} libraries.
To carry implement the flood sequence generating and estimating functions, a new package called \texttt{floodsampling} was developed and is available at \url{github.com/jdossgollin/floodsampling}.
The codes used to generate the figures and text of this paper are available at \url{github.com/jdossgollin/MNpaper}.

@@

\begin{figure}[ht]
  \caption{50 streamflow sequences from (Top) the ENSO-based model described in \cref{methods-enso} and (Bottom) the Markov chain modle described in \cref{methods-markov} with \(\gamma=0\), implying stationarity\label{fig:stationary-sequences}}
\end{figure}

\begin{figure}[ht]
  \caption{50 streamflow sequences from (Top) the ENSO-based model described in \cref{methods-enso} and (Bottom) the Markov chain modle described in \cref{methods-markov}. To add secular variability, \(\gamma=0.01\) for the ENSO-based model, implying a linear trend, and for the Markov model \(\gamma_1 = 0.01\), implying a linear trend only in the ``wet'' state.\label{fig:trend-sequences}}
\end{figure}

@@

\begin{figure}[ht]
  \caption{The estimation bias for 1000* sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \caption{The estimation variance for 1000* sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

@@

\begin{figure}[ht]
  \caption{The estimation bias for 1000* sequences generated by each of two non-stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \caption{The estimation variance for 1000* sequences generated by each of two non-stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

@@

We still need to find a good case study; it can be a highly simplified one as long as there is a long streamflow record available and building levees is an option that has been proposed.
If there's not an obvious candidate, we can use one of our models above.
We can make some other strong assumptions
\begin{itemize}
  \item Risk exposure constant in time, which allows us to focus on the estimation rather than comprehensive risk manageement
  \item Constant future discounting rate
  \item Flood damages are linearly related to streamflow over some amount:
  \begin{equation}
    \mathcal{D}(t) = k \qty( Q(t) - Q_T)
  \end{equation}
\end{itemize}
Then our two options are to build a levee, which we can assume to have a capital cost plus operation and maintenance cost which is proportional to the capital cost, which increases \(Q_T\) by some fixed amount, or to buy insurance.
For the insurance, we consider a contract in which an insurer agrees to pay the entire coverage cost \(C\) every time that the trigger threshold \(Q_T\) is exceeded.
Although \(Q_T\) is in general a decision variable, we treat it as fixed.
Then, using the posterior distribution of the parameters of the chosen statistical model (from \cref{tab:model-fitting}) we estimate the probability that \(Q(t) \geq Q_T\) for \(t \in \vb{t}\), which is equivalent to \(p_T(t)\).
For a particular year, the fair price for the insurance premium is \(\expectation \qty(p_T) C\).
However, there is typically also a risk premium associated with the policy that prices the uncertainty in this estimate (since the trigger \(Q_T\) is specified in the contract and thus not uncertain).
This risk premium is proportional to the variance of \(p_T\), or more generally to its uncertainty distribution.
The insurance premium for a particular year is thus
\begin{equation} \label{eq:premium}
  Z(t) = C \qty[\expectation \qty(p_T(t)) + R \variance \qty(p_T(t))]
\end{equation}
where \(R \geq 0\) is a coefficient describing the cost of tue uncertainty in the estimate of \(p_T(t)\).

Then at each point in time, calculate:
\begin{itemize}
  \item Expected cost plus damage per year for the levee using perfect knowledge of the next 101 years
  \item Expected cost plus damage per year for the levee using MCMC projections of the next 101 years from all 3 models
  \item Expected cost plus damage per year for a 5-year insurance contract using perfect knowledge of the next 5 years
  \item Expected cost plus damage per year for a 5-year insurance contract using MCMC projections of the next 5 years from all 3 models
\end{itemize}

Can add appropriate plot or plots.

@@

From a decision making perspective, one could either design and build a project that gives us protection through the year 2100 considering both types of nonstationarities, or consider a sequence of projects every M years that incrementally add (or not) protection. How does the uncertainty and hence risk for these decisions manifest depending on the design level p, the duration M, and the uncertainty associated with the estimation of the \( f(Qp(t)) \) given a methodology used for prediction -- \eg{} the model chain using gcms and such and models fitted to a historical data length \(M\).
WHat is the implication for choices of p and M under different models for stationarity and uncertainty using some simple loss and cost  functions.

How could  financial risk mitigation instruments be used in conjunction with structural design to identify a robust, adaptive climate risk mitigation strategy with sequential decisions that are informed by updated estimates of risk.

@@

% -----------------------------------------------------------------------------
% END HERE
% -----------------------------------------------------------------------------
\appendix
\printbibliography{}

\section{Notation}

@@

\section{Supplemental Figures}

\begin{figure}
  \caption{Insert here a 1000-year plot of the ENSO time series\label{fig:enso-ts}}
\end{figure}

\begin{figure}
  \caption{Insert here a global wavelet or spectral plot of the ENSO time series\label{fig:enso-spectrum}}
\end{figure}
